---
title: "Case Studies"
subtitle: "Lecture 6 - LAS"
author: "Don van den Bergh"
date: 2026-09-28
date-format: "DD-MM-YYYY"
format:
    dtslides-revealjs:
        theme: [../common_resources/css/tilburg-university.scss]
        logo: ../common_resources/logos/logo_tu_small.svg
        slide-number: c/t
        toc: true
        toc-depth: 1
        toc-title: "Today"
        embed-resources: true
---

## Today {.unlisted}

```{r}
#| include: false
#| label: setup

library(dplyr)
library(ggplot2)
library(DT)
library(kableExtra)

options(knitr.kable.NA = "")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

set.seed(2026)

# ------------------------------------------------------------
# TODO BEFORE THE LECTURE
# ------------------------------------------------------------
# 1. Check that the Bechdel data path below matches your project.
# 2. Decide how to collect the live sleep data (Mentimeter / Forms /
#    show of hands / typed values) and replace the placeholder vector.
# 3. Consider downloading billboard.csv once to scripts/billboard.csv
#    if you want the lecture to render completely offline.
# 4. Render once and check that the fixed random samples give results
#    that are pedagogically useful.
# 5. Optional: replace some tables by screenshots / images if preferred.
# ------------------------------------------------------------

# Useful plotting theme for this lecture
theme_set(
  theme_minimal(base_size = 18) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot"
    )
)

# A small helper for displaying p-values
pretty_p <- function(p) {
  if (p < .001) "< .001" else paste0("= ", format(round(p, 3), nsmall = 3))
}
```

$$
\newcommand{\hypo}{\mathcal{H}}
$$



Two statistical tests:

- Z-test
- $t$-test

These are mostly the names for tests we've already applied during the previous lectures.

. . .

Three very different datasets.

One recurring workflow.


# The Statistical Workflow {.section-slide}


## From a Question to a Conclusion

::: {style="font-size: 90%;"}

1. What is the _research question_?
2. What are the _population_, _sample_, _variables_, and _measurement levels_?
3. What do the _raw data_ look like?
4. Which _descriptive statistics_ are useful?
5. What are $\hypo_0$ and $\hypo_1$?
6. What would we expect because of _sampling variability_ if $\hypo_0$ were true?
7. How far is our estimate from $\hypo_0$ in _standard errors_?
8. What is the _$p$-value_?
9. What does a _confidence interval_ tell us?
10. What _can_ and what _can't_ we conclude?

:::


## One Quantity We Will Keep Using

For a sample mean,

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}.
$$

The generic test statistic is

$$
\text{test statistic}
=
\frac{\text{observed estimate} - \text{value under }\hypo_0}
     {\text{standard error}}.
$$

. . .

Interpretation:

_How many standard errors away from the null hypothesis is our estimate?_


<!-- ## Before We Start -->

<!-- TODO: Insert QR code / Mentimeter link here. -->

<!-- One question: -->

<!-- > **How many hours did you sleep last night?** -->

<!-- Please answer now. -->

<!-- . . . -->

<!-- We will come back to your answers later. -->


# Case Study 1: Hit Songs {.section-slide}


## How Long Is a Hit Song?

People often say:

> "Pop songs are about three minutes long."

Is that a reasonable description of songs that reach number 1 on the Billboard Hot 100?


::: footer
<https://en.wikipedia.org/wiki/Three-minute_pop_song>
:::

## The Data

Let's use a dataset on the Billboard Hot 100 Number Ones.

It contains songs that reached number 1 between 1958 and 2025, including their length in seconds.

```{r}
#| include: false
# Source:
# https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-26
#
# TODO: If you want fully offline rendering, download this file once
#       and replace the URL by "scripts/billboard.csv".

# billboard_url <- paste0(
#   "https://raw.githubusercontent.com/rfordatascience/tidytuesday/",
#   "main/data/2025/2025-08-26/billboard.csv"
# )
billboard_url <- "data/billboard.csv"


billboard <- read.csv(billboard_url)

billboard_population <- billboard |>
  filter(!is.na(length_sec), length_sec > 0)

# We deliberately pretend that we only observed a random sample.
set.seed(1209)
billboard_sample <- billboard_population |>
  slice_sample(n = 30)
```

::: footer
Data: Chris Dalla Riva / TidyTuesday, Billboard Hot 100 Number Ones.
Source: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-26>
:::


## Population and Sample

What is the population we are trying to say something about?

. . .

- Billboard Hot 100 number-one songs, 1958–2025.

. . .

What is the measurement level?

. . .

- Ratio, because distances have a meaning and zero has a meaning.

. . .

For today's analysis:

- _Population_: Billboard Hot 100 number-one songs in this database
- _Sample_: 30 randomly selected number-one songs
- _Variable_: song length in seconds
- _Measurement level_: ratio


## Our Sample

```{r}
#| echo: false

billboard_sample |>
  transmute(
    Song = song,
    Artist = artist,
    Year = format(as.Date(date), "%Y"),
    `Length (seconds)` = length_sec
  ) |>
  head(12) |>
  DT::datatable(
    options = list(
      searching = FALSE,
      info = FALSE,
      paging = FALSE,
      scrollX = TRUE,
      scrollY = "360px",
      scrollCollapse = TRUE
    ),
    rownames = FALSE
  )
```


## First: Look at the Data

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(billboard_sample, aes(x = length_sec)) +
  geom_histogram(binwidth = 20, boundary = 0) +
  geom_vline(xintercept = 180, linetype = 2, linewidth = 1) +
  labs(
    x = "Song length (seconds)",
    y = "Number of songs",
    title = "Lengths of the 30 sampled number-one hits"
  )
```

What do you notice?


## Describe the Sample

```{r}
#| include: false

song_n <- nrow(billboard_sample)
song_mean <- mean(billboard_sample$length_sec)
song_median <- median(billboard_sample$length_sec)
song_sd <- sd(billboard_sample$length_sec)
song_se <- song_sd / sqrt(song_n)
```

::: {style="font-size: 105%;"}

- $n = `{r} song_n`$
- Mean = `{r} round(song_mean, 1)` seconds
- Median = `{r} round(song_median, 1)` seconds
- SD = `{r} round(song_sd, 1)` seconds

:::

. . .

The median is near the mean, what does this tell us?

. . .

Distribution is roughly symmetric / little skew.

. . .

The sample mean is `{r} round(song_mean - 180, 1)` seconds away from three minutes.

Is that a large difference?

## Hypotheses

Let $\mu$ be the population mean song length.

We test:

$$
\hypo_0: \mu = 180 \qquad \text{against} \qquad \hypo_1: \mu \neq 180.
$$

Is this one or two tailed?

. . .

Two-tailed. Why?


## Assume the Null Hypothesis is true

If $\hypo_0$ is true, the population mean is 180 seconds.

Even then, a random sample of 30 songs will almost never have a mean of *exactly* 180.

. . .

Why?

Sampling variability!


## Standard Error

Our estimate of the standard error is

$$
SE_{\bar{x}}
=
\frac{s}{\sqrt{n}}.
$$

For our data,

$$
SE_{\bar{x}}
=
\frac{`r round(song_sd, 1)`}
{\sqrt{`r song_n`}}
=
`r round(song_se, 2)` \text{ seconds}.
$$

What does this number represent?

. . .

Note that we _estimate_ SE rather than that we know it!

## Test Statistic

```{r}
#| include: false

song_t <- (song_mean - 180) / song_se
song_df <- song_n - 1
song_p <- 2 * pt(-abs(song_t), df = song_df)
```

$$
t =\frac{\bar{x} - 180}
     {SE_{\bar{x}}}
=\frac{`r round(song_mean, 1)` - 180}
     {`r round(song_se, 2)`}
= `r round(song_t, 2)`.
$$

. . .

Our sample mean lies about `{r} abs(round(song_t, 2))` standard errors from the value predicted by $\hypo_0$.

The "degrees of freedom" are $df = 29$.

## The Sampling Distribution Under $\hypo_0$

```{r}
#| fig-width: 9
#| fig-height: 5

x_t <- seq(-4.5, 4.5, length.out = 1000)
dens_t <- dt(x_t, df = song_df)
t_plot <- data.frame(t = x_t, density = dens_t)

ggplot(t_plot, aes(t, density)) +
  geom_line(linewidth = 1.2) +
  geom_area(
    data = subset(t_plot, abs(t) >= abs(song_t)),
    aes(y = density),
    alpha = .35
  ) +
  geom_vline(xintercept = c(-abs(song_t), abs(song_t)), linetype = 2) +
  labs(
    x = "Test statistic under H0",
    y = "Density",
    title = "How unusual is our observed result if H0 is true?"
  )
```


## The $p$-value

We know $df = n - 1 = 30 - 1 = 29$ and $t = 5.26$

. . .

Let's examine the $t$-table to find the $p$-value.

. . .

Look up the row matching $df = 29$ and go right until the next value is larger than $t=5.26$.

. . .

The probability of a result at least this extreme under $\hypo_0$ is

$$
p < 0.001.
$$

Using $\alpha = .05$, we reject $\mathcal{H}_0$.

. . .

Why don't we need to know the exact $p$-value?

## Confidence Interval

```{r}
#| include: false

song_crit <- qt(.975, df = song_df)
song_ci <- song_mean + c(-1, 1) * song_crit * song_se
```

A 95% confidence interval for the population mean is

$$
[
\bar{x} - t_{95\%} \times SE_{\bar{x}},\;
\bar{x} + t_{95\%} \times SE_{\bar{x}}
]
$$

. . .

$$
[`r round(song_ci[1], 1)`,\; `r round(song_ci[2], 1)`]
\text{ seconds}.
$$

. . .

Does 180 seconds fall inside this interval?

How does that relate to our hypothesis test?

. . .

::: {style="font-size: 80%;"}

- Because $\sigma$ is unknown, we replaced it with $s$. That means our standardized statistic follows a $t$-distribution rather than a standard normal distribution.
- Compared to last week, the only difference is using a $t$ distribution instead of a $Z$-distribution for the $p$-value and confidence interval.

:::

## What if we examine all the data?

For teaching purposes, I pretended that these 30 songs were all data we had.

. . .

But this dataset contains 1177 songs obtained between 1958 and 2025.

So now we can look behind the curtain.


## The Population Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

population_mean_song <- mean(billboard_population$length_sec)
population_sd_song <- sd(billboard_population$length_sec)

ggplot(billboard_population, aes(x = length_sec)) +
  geom_histogram(binwidth = 15, boundary = 0) +
  geom_vline(xintercept = population_mean_song, linewidth = 1) +
  geom_vline(xintercept = 180, linetype = 2, linewidth = 1) +
  labs(
    x = "Song length (seconds)",
    y = "Number of songs",
    title = "Population distribution of number-one songs"
  )
```

- Population mean: `{r} round(population_mean_song, 1)` seconds.
- Sample mean: `{r} round(song_mean, 1)` seconds.

## Case Study 1: Summary

::: {style="font-size: 90%;"}

- Population and sample
- Measurement levels
- Histograms
- Mean, median, and standard deviation
- Hypotheses
- Sampling variability
- Standard error
- Sampling distribution
- Test statistic
- $p$-value
- Confidence interval

:::

## Case Study 1: Summary

We use a one-sample $t$-test because the standard devation was unknown/ estimated from the data.

What if we do know the standard deviation?


# Case Study 2: Movies {.section-slide}

## Does Passing the Bechdel Test Pay?

Recall the Bechdel test.

A movie passes if:

1. it contains at least two named women,
2. who talk to each other,
3. about something other than a man.

. . .

Now we ask a different question:

> Among movies that pass, does the average movie earn back its production budget?


## Operationalize "Earn Back Its Budget"

To turn that question into a variable, let's define _return on investment_:

$$
ROI
=
\frac{\text{worldwide gross}}
     {\text{production budget}}.
$$

Then:

- $ROI = 1$: worldwide gross equals the production budget
- $ROI > 1$: worldwide gross exceeds the production budget
- $ROI < 1$: worldwide gross is below the production budget

. . .

Is this the same thing as _profit_?


## Operationalize "Earn Back Its Budget"

No.

Production budget is not the movie's complete cost.

Worldwide box office is not the amount that goes directly to the studio.

. . .

So technically our variable is better described as:

"Worldwide box-office gross relative to production budget"

This is an _operationalization_ of financial performance.


```{r}
#| include: false

bechdel <- read.csv("../lecture2/scripts/movies.csv")

# The FiveThirtyEight dataset normally has columns named:
# binary, budget, intgross
bechdel_roi <- bechdel |>
  filter(
    binary == "PASS",
    !is.na(budget),
    !is.na(intgross),
    budget > 0,
    intgross > 0
  ) |>
  mutate(
    intgross = suppressWarnings(as.numeric(intgross)),
    roi = intgross / budget
  )

# We again pretend that we observed only a random sample.
set.seed(2409)
bechdel_sample <- bechdel_roi |>
  slice_sample(n = min(50, nrow(bechdel_roi)))
```


## Before Calculating Anything...

What do you expect the distribution of ROI to look like?

. . .

- Symmetric?
- Left-skewed?
- Right-skewed?
- Any outliers?


## Look at the Raw Data

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(bechdel_sample, aes(x = roi)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 1, linetype = 2, linewidth = 1) +
  labs(
    x = "Worldwide gross / production budget",
    y = "Number of movies",
    title = "ROI among sampled movies that pass the Bechdel test"
  )
```

What do you notice?


## Mean or Median?

```{r}
#| include: false

movie_n <- nrow(bechdel_sample)
movie_mean <- mean(bechdel_sample$roi)
movie_median <- median(bechdel_sample$roi)
movie_sd <- sd(bechdel_sample$roi)
movie_sigma <- sd(bechdel_roi$roi)
movie_se <- movie_sigma / sqrt(movie_n)
```

::: {style="font-size: 105%;"}

- Mean ROI = `{r} round(movie_mean, 2)`
- Median ROI = `{r} round(movie_median, 2)`
- SD = `{r} round(movie_sd, 2)`

:::

. . .

Why can the mean and median differ substantially here?


## But Our Hypothesis Is About a Mean

Let $\mu$ be the _population mean ROI_ among passing movies in our sampling frame.

We ask whether the average exceeds break-even:

$$
\hypo_0: \mu = 1
$$

against

$$
\hypo_1: \mu > 1.
$$

. . .

This is a _one-tailed Z-test_.


## Your Turn: What Comes Next?

<!-- Give students 1-2 minutes before revealing. -->

We know:

- $\bar{x} = `{r} round(movie_mean, 2)`$
- $n = `{r} movie_n`$

We assume: $\sigma = `{r} round(movie_sigma, 2)`$ as known from the entire dataset.

- Typically there is some previous large-scale industry research that has established that the population standard deviation of this gross-to-budget ratio is $\sigma = `{r} round(movie_sigma, 2)`$.

What should we calculate next?

. . .

$$
SE_{\bar{x}} = \frac{\sigma}{\sqrt{n}}.
$$


## Standard Error

$$
SE_{\bar{x}}
=
\frac{`r round(movie_sigma, 2)`}
{\sqrt{`r movie_n`}}
=
`r round(movie_se, 3)`.
$$

. . .

Notice:

> A highly variable raw distribution can still give a much less variable distribution of **sample means**.


## Your Turn: Test Statistic

Under $\hypo_0$, the expected population mean is 1.

What is the test statistic?

. . .

```{r}
#| include: false

movie_z <- (movie_mean - 1) / movie_se
movie_p <- 1 - pnorm(movie_z)
```

$$
Z
=
\frac{`r round(movie_mean, 2)` - 1}
     {`r round(movie_se, 3)`}
=
`r round(movie_z, 2)`.
$$


## Interpret It Before Looking at $p$

Our observed mean is `{r} abs(round(movie_z, 2))` standard errors above the value predicted by $\hypo_0$.

. . .

If you had to guess the $p$-value, is it:

- large,
- around .05,
- or very small?


## Sampling Distribution and $p$-value

```{r}
#| fig-width: 9
#| fig-height: 5

movie_x_z <- seq(-4, max(6, movie_z + 1), length.out = 1200)
movie_z_plot <- data.frame(
  z = movie_x_z,
  density = dnorm(movie_x_z)
)

ggplot(movie_z_plot, aes(z, density)) +
  geom_line(linewidth = 1.2) +
  geom_area(
    data = subset(movie_z_plot, z >= movie_z),
    aes(y = density),
    alpha = .35
  ) +
  geom_vline(xintercept = movie_z, linetype = 2) +
  labs(
    x = "Test statistic under H0",
    y = "Density",
    title = "One-tailed probability under the null hypothesis"
  )
```


## Conclusion

$$
p = `{r} format(movie_p, digits = 2)`.
$$

Using $\alpha=.05$:

```{r}
#| results: asis
if (movie_p < .05) {
  cat("**We reject $\\mathcal{H}_0$.**")
} else {
  cat("**We do not reject $\\mathcal{H}_0$.**")
}
```



## Confidence Interval

```{r}
#| include: false

movie_crit <- qnorm(.975)
movie_ci <- movie_mean + c(-1, 1) * movie_crit * movie_se
```

The ordinary two-sided 95% confidence interval is

$$
[`r round(movie_ci[1], 2)`,\; `r round(movie_ci[2], 2)`].
$$

What does this add beyond the $p$-value?

. . .

It contains 0!

. . .

It contains ~~0~~1! The value expected under $\hypo_0$ is 1!

## Does Skewness Break Everything?

The **raw ROI values** are strongly skewed.

But our test concerns the **sample mean**.

So the relevant question is:

> What does the sampling distribution of the sample mean look like?


## Simulate the Sampling Distribution

```{r}
#| include: false

set.seed(2410)
B <- 5000

sim_movie_means <- replicate(
  B,
  mean(sample(bechdel_roi$roi, size = movie_n, replace = TRUE))
)

sim_movies <- data.frame(mean = sim_movie_means)
```


## Raw Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(bechdel_roi, aes(x = roi)) +
  geom_histogram(bins = 50) +
  labs(
    x = "ROI",
    y = "Number of movies",
    title = "Individual movies: strongly right-skewed"
  )
```


## Sampling Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(sim_movies, aes(x = mean)) +
  geom_histogram(bins = 40) +
  labs(
    x = paste0("Mean ROI in a sample of ", movie_n, " movies"),
    y = "Number of simulated samples",
    title = "Means of repeated samples"
  )
```

. . .

The Central Limit Theorem is about the _second_ distribution.

## Sampling Distribution

In this case, the sampling distribution doesn't look normal.

Keep in mind, that the CLT talks about _the limit_.

Typically $n = 30$ suffices, in this (extreme) case it did not.

## What if we had larger samples?


```{r}
#| fig-width: 9
#| fig-height: 5

set.seed(2409)
ns <- c(50, 100, 200, 400, 600, 1000)
B <- 5000

sim_movies_tmp <- data.frame(
  mean = numeric(B),
  n = numeric(B)
)
sim_movies2 <- NULL

for (i in 1:6) {

  sim_movie_means_tmp <- replicate(B,
    mean(sample(bechdel_roi$roi, size = ns[i], replace = TRUE))
  )
  sim_movies_tmp$mean <- sim_movie_means_tmp
  sim_movies_tmp$n <- ns[i]
  sim_movies2 <- rbind(sim_movies2, sim_movies_tmp)
}

ggplot(sim_movies2, aes(x = mean)) +
  geom_histogram(bins = 40) +
  facet_wrap(. ~ n, labeller = label_both) +
  labs(
    x = paste0("Mean ROI in a sample of ", movie_n, " movies"),
    y = "Number of simulated samples",
    title = "Means of repeated samples"
  )
```


## A More Important Problem

Suppose our calculations are all correct.

What exactly can we generalize to?

. . .

The dataset is _not_ a simple random sample of every movie ever made.


## Statistical Significance Is Not Enough

A small $p$-value answers a narrow question:

- Are these data evidence against $\hypo_0$?

It does _not_ automatically tell us:

- whether the effect is important,
- whether the measurement is ideal,
- whether the sample represents the target population


## Case Study 2: What Changed?

The statistical machinery was nearly identical.

- We used a $Z$-test instead of a $t$-test.

But the data raised new questions about:

- skewness and outliers,
- mean versus median,
- operationalization,
- the target population,
- external validity.

. . .

Good statistics is more than calculating the correct formula.

## The Core Pattern

In both  cases:

$$
\text{observed sample}
\longrightarrow
\text{estimate}
\longrightarrow
\text{standard error}
$$

$$
\longrightarrow
\text{sampling distribution under }\hypo_0
\longrightarrow
p
\longrightarrow
\text{conclusion}.
$$

The **story** changes.

The statistical logic does not.


## What Does a Test Statistic Mean?

Suppose

$$
t = -2.4.
$$

What does that mean?

. . .

> The observed estimate lies **2.4 standard errors below** the value predicted by the null hypothesis.

This interpretation is much more useful than memorizing formulas.


## What Does a $p$-value Mean?

Suppose $p=.03$.

Which statement is correct?

A. There is a 3% probability that $\hypo_0$ is true.

B. There is a 3% probability that the result occurred by chance.

C. If $\hypo_0$ were true, results at least this extreme would occur about 3% of the time.

D. There is a 97% probability that $\hypo_1$ is true.

. . .

**C**


## What Does a Confidence Interval Mean?

A 95% confidence interval is **not**:

> "There is a 95% probability that $\mu$ lies in this particular interval."

Instead:

> If we repeated the procedure many times, 95% of the intervals produced by this procedure would contain the true population value.

. . .

(The first interpretation is the Bayesian credible interval)

## Raw Distribution vs. Sampling Distribution

This distinction appeared repeatedly today.

_Raw-data distribution_

- describes individual observations
- e.g., individual movies or songs

_Sampling distribution_

- describes a statistic over repeated samples
- e.g., sample means
- determines the uncertainty of our estimate


## What Happens When $n$ Increases?

If the population variability stays the same:

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}.
$$

As $n$ increases:

- the standard error decreases,
- the sampling distribution becomes narrower,
- estimates become more precise,
- power generally increases.


## But Large $n$ Does Not Fix Everything

A huge sample does _not_ automatically fix:

- poor measurement,
- selection bias,
- an irrelevant target population,
- confounding,
- bad research questions.

. . .

Statistics quantifies _sampling uncertainty_.

## Comparison

| | Billboard | Bechdel |
|---|---:|---:|
| Question about | Mean | Mean |
| $\sigma$ | Unknown | Known |
| Standard error | $s/\sqrt{n}$ | $\sigma/\sqrt{n}$ |
| Test statistic | $t$ | $Z$ |
| Reference distribution | $t_{n-1}$ | $N(0,1)$ |
| $df$ | $n-1$ | — |
| Basic logic | Same | Same |
