---
title: "Case Studies"
subtitle: "Lecture 6 - LAS"
author: "Don van den Bergh"
date: 2026-09-28
date-format: "DD-MM-YYYY"
format:
    dtslides-revealjs:
        theme: [../common_resources/css/tilburg-university.scss]
        logo: ../common_resources/logos/logo_tu_small.svg
        slide-number: c/t
        toc: true
        toc-depth: 1
        toc-title: "Today"
        embed-resources: true
---

```{r}
#| include: false
#| label: setup

library(dplyr)
library(ggplot2)
library(DT)
library(kableExtra)

options(knitr.kable.NA = "")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

set.seed(2026)

# ------------------------------------------------------------
# TODO BEFORE THE LECTURE
# ------------------------------------------------------------
# 1. Check that the Bechdel data path below matches your project.
# 2. Decide how to collect the live sleep data (Mentimeter / Forms /
#    show of hands / typed values) and replace the placeholder vector.
# 3. Consider downloading billboard.csv once to scripts/billboard.csv
#    if you want the lecture to render completely offline.
# 4. Render once and check that the fixed random samples give results
#    that are pedagogically useful.
# 5. Optional: replace some tables by screenshots / images if preferred.
# ------------------------------------------------------------

# Useful plotting theme for this lecture
theme_set(
  theme_minimal(base_size = 18) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot"
    )
)

# A small helper for displaying p-values
pretty_p <- function(p) {
  if (p < .001) "< .001" else paste0("= ", format(round(p, 3), nsmall = 3))
}
```

$$
\newcommand{\hypo}{\mathcal{H}}
$$


## Today {.unlisted}

- Z-test
- $t$-test

These are mostly the names for tests we've already applied during the previous lectures.

. . .

Three very different datasets.

One recurring workflow.


# The Statistical Workflow {.section-slide}


## From a Question to a Conclusion

::: {style="font-size: 90%;"}

1. What is the _research question_?
2. What are the _population_, _sample_, _variables_, and _measurement levels_?
3. What do the _raw data_ look like?
4. Which _descriptive statistics_ are useful?
5. What are $\hypo_0$ and $\hypo_1$?
6. What would we expect because of __sampling variability_ if $\hypo_0$ were true?
7. How far is our estimate from $\hypo_0$ in _standard errors_?
8. What is the _$p$-value_?
9. What does a _confidence interval_ tell us?
10. What _can_ and what _can't_ we conclude?

:::


## One Quantity We Will Keep Using

For a sample mean,

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}.
$$

The generic test statistic is

$$
\text{test statistic}
=
\frac{\text{observed estimate} - \text{value under }\hypo_0}
     {\text{standard error}}.
$$

. . .

Interpretation:

_How many standard errors away from the null hypothesis is our estimate?_


<!-- ## Before We Start -->

<!-- TODO: Insert QR code / Mentimeter link here. -->

<!-- One question: -->

<!-- > **How many hours did you sleep last night?** -->

<!-- Please answer now. -->

<!-- . . . -->

<!-- We will come back to your answers later. -->


# Case Study 1: Hit Songs {.section-slide}


## How Long Is a Hit Song?

People often say:

> "Pop songs are about three minutes long."

Is that a reasonable description of songs that reach number 1 on the Billboard Hot 100?


::: footer
<https://en.wikipedia.org/wiki/Three-minute_pop_song>
:::

## The Data

Let's use a dataset on the Billboard Hot 100 Number Ones.

It contains songs that reached number 1 between 1958 and 2025, including their length in seconds.

```{r}
#| include: false
# Source:
# https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-26
#
# TODO: If you want fully offline rendering, download this file once
#       and replace the URL by "scripts/billboard.csv".

billboard_url <- paste0(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/",
  "main/data/2025/2025-08-26/billboard.csv"
)

billboard <- read.csv(billboard_url)

billboard_population <- billboard |>
  filter(!is.na(length_sec), length_sec > 0)

# We deliberately pretend that we only observed a random sample.
set.seed(1209)
billboard_sample <- billboard_population |>
  slice_sample(n = 30)
```

::: footer
Data: Chris Dalla Riva / TidyTuesday, Billboard Hot 100 Number Ones.
Source: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-26>
:::


## Population and Sample

What is the population we are trying to say something about?

. . .

Popular songs

. . .

What is the measurement level?

. . .

Ratio, because distances have a meaning and zero has a meaning.

. . .

For today's analysis:

- _Population_: Billboard Hot 100 number-one songs in this database
- _Sample_: 30 randomly selected number-one songs
- _Variable_: song length in seconds
- _Measurement level_: ratio


## Our Sample

```{r}
#| echo: false

billboard_sample |>
  transmute(
    Song = song,
    Artist = artist,
    Year = format(as.Date(date), "%Y"),
    `Length (seconds)` = length_sec
  ) |>
  head(12) |>
  DT::datatable(
    options = list(
      searching = FALSE,
      info = FALSE,
      paging = FALSE,
      scrollX = TRUE,
      scrollY = "360px",
      scrollCollapse = TRUE
    ),
    rownames = FALSE
  )
```


## First: Look at the Data

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(billboard_sample, aes(x = length_sec)) +
  geom_histogram(binwidth = 20, boundary = 0) +
  geom_vline(xintercept = 180, linetype = 2, linewidth = 1) +
  labs(
    x = "Song length (seconds)",
    y = "Number of songs",
    title = "Lengths of the 40 sampled number-one hits"
  )
```

What do you notice?


## Describe the Sample

```{r}
#| include: false

song_n <- nrow(billboard_sample)
song_mean <- mean(billboard_sample$length_sec)
song_median <- median(billboard_sample$length_sec)
song_sd <- sd(billboard_sample$length_sec)
song_se <- song_sd / sqrt(song_n)
```

::: {style="font-size: 105%;"}

- $n = `{r} song_n`$
- Mean = `{r} round(song_mean, 1)` seconds
- Median = `{r} round(song_median, 1)` seconds
- SD = `{r} round(song_sd, 1)` seconds

:::

. . .

The sample mean is `{r} round(song_mean - 180, 1)` seconds away from three minutes.

Is that a **large** difference?

. . .

The median is near the mean, what does this tell us?

. . .

Distribution is roughly symmetric / little skew.

## Research Hypotheses

Let $\mu$ be the population mean song length.

We test:

$$
\hypo_0: \mu = 180 \qquad \text{against} \qquad \hypo_1: \mu \neq 180.
$$

Is this one or two tailed?

. . .

Two-tailed. Why?


## Assume the Null Hypothesis

If $\hypo_0$ is true, the population mean is 180 seconds.

Even then, a random sample of 30 songs will almost never have a mean of *exactly* 180.

. . .

Why?

Sampling variability!


## Standard Error

Our estimate of the standard error is

$$
SE_{\bar{x}}
=
\frac{s}{\sqrt{n}}.
$$

For our data,

$$
SE_{\bar{x}}
=
\frac{`r round(song_sd, 1)`}
{\sqrt{`r song_n`}}
=
`r round(song_se, 2)` \text{ seconds}.
$$

What does this number represent?


## Test Statistic

```{r}
#| include: false

song_t <- (song_mean - 180) / song_se
song_df <- song_n - 1
song_p <- 2 * pt(-abs(song_t), df = song_df)
```

$$
t
=
\frac{\bar{x} - 180}
     {SE_{\bar{x}}}
=
\frac{`r round(song_mean, 1)` - 180}
     {`r round(song_se, 2)`}
=
`r round(song_t, 2)`.
$$

. . .

Our sample mean lies about `{r} abs(round(song_t, 2))` standard errors from the value predicted by $\hypo_0$.

The "degrees of freedom" are $df = 29$.


## The Sampling Distribution Under $\hypo_0$

```{r}
#| fig-width: 9
#| fig-height: 5

x_t <- seq(-4.5, 4.5, length.out = 1000)
dens_t <- dt(x_t, df = song_df)
t_plot <- data.frame(t = x_t, density = dens_t)

ggplot(t_plot, aes(t, density)) +
  geom_line(linewidth = 1.2) +
  geom_area(
    data = subset(t_plot, abs(t) >= abs(song_t)),
    aes(y = density),
    alpha = .35
  ) +
  geom_vline(xintercept = c(-abs(song_t), abs(song_t)), linetype = 2) +
  labs(
    x = "Test statistic under H0",
    y = "Density",
    title = "How unusual is our observed result if H0 is true?"
  )
```


## The $p$-value

. . .

Examine the $t$-table.

. . .

Look up the row matching $df = 29$ and go right until the next value is larger than $t$.

. . .

The probability of a result at least this extreme under $\hypo_0$ is

$$
p < 0.001.
$$

Using $\alpha = .05$:

We reject $\mathcal{H}_0$.



## Confidence Interval

```{r}
#| include: false

song_crit <- qt(.975, df = song_df)
song_ci <- song_mean + c(-1, 1) * song_crit * song_se
```

A 95% confidence interval for the population mean is

$$
[`r round(song_ci[1], 1)`,\; `r round(song_ci[2], 1)`]
\text{ seconds}.
$$

. . .

Does 180 seconds fall inside this interval?

How does that relate to our hypothesis test?


## We Have Been Cheating

For teaching purposes, we pretended that these 30 songs were all the data we had.

. . .

But this dataset contains **the whole sampling frame**.

So now we can look behind the curtain.


## The Population Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

population_mean_song <- mean(billboard_population$length_sec)
population_sd_song <- sd(billboard_population$length_sec)

ggplot(billboard_population, aes(x = length_sec)) +
  geom_histogram(binwidth = 15, boundary = 0) +
  geom_vline(xintercept = population_mean_song, linewidth = 1) +
  geom_vline(xintercept = 180, linetype = 2, linewidth = 1) +
  labs(
    x = "Song length (seconds)",
    y = "Number of songs",
    title = "Population distribution of number-one songs"
  )
```

Population mean: **`{r} round(population_mean_song, 1)` seconds**.


## Raw Data vs. Sampling Distribution

These are different distributions:

1. Distribution of **individual song lengths**
2. Distribution of **sample means**

Let's simulate the second one.


## Simulate Repeated Samples

```{r}
#| include: false

set.seed(1210)
B <- 5000
sim_song_means <- replicate(
  B,
  mean(sample(billboard_population$length_sec, size = song_n, replace = TRUE))
)

sim_song <- data.frame(mean = sim_song_means)
```


## Sampling Distribution of the Mean

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(sim_song, aes(x = mean)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = population_mean_song, linewidth = 1) +
  labs(
    x = "Mean song length in a sample of 30",
    y = "Number of simulated samples",
    title = "5,000 hypothetical repeated samples"
  )
```

What happened to the shape and spread?


## The Central Limit Theorem, in Action

The **raw population** can be somewhat irregular.

The distribution of **sample means** is much more regular.

. . .

Its standard deviation should be approximately

$$
\frac{\sigma}{\sqrt{n}}.
$$

```{r}
#| include: false

empirical_se_song <- sd(sim_song_means)
theoretical_se_song <- population_sd_song / sqrt(song_n)
```

Simulation: `{r} round(empirical_se_song, 2)`

Formula: `{r} round(theoretical_se_song, 2)`


## Case Study 1: What Did We Use?

::: {style="font-size: 90%;"}

- Population and sample
- Measurement levels
- Histograms
- Mean, median, and standard deviation
- Hypotheses
- Sampling variability
- Standard error
- Sampling distribution
- Test statistic
- $p$-value
- Confidence interval
- Central Limit Theorem

:::

. . .

Almost the entire first half of the course from **one question**.


# Case Study 2: Movies {.section-slide}


## Does Passing the Bechdel Test Pay?

Recall the Bechdel test.

A movie passes if:

1. it contains at least two named women,
2. who talk to each other,
3. about something other than a man.

. . .

Now we ask a different question:

> Among movies that pass, does the average movie earn back its production budget?


## Operationalize "Earn Back Its Budget"

We need to turn that question into a variable.

Define

$$
ROI
=
\frac{\text{worldwide gross}}
     {\text{production budget}}.
$$

Then:

- $ROI = 1$: worldwide gross equals the production budget
- $ROI > 1$: worldwide gross exceeds the production budget
- $ROI < 1$: worldwide gross is below the production budget

. . .

Is this literally the same thing as **profit**?


## Important Measurement Caveat

No.

Production budget is not the movie's complete cost.

Worldwide box office is not the amount that goes directly to the studio.

. . .

So our variable is better described as:

> **Worldwide box-office gross relative to production budget**

This is an **operationalization** of financial performance.


## Load the Bechdel Data

```{r}
#| include: false

bechdel <- read.csv("../lecture2/scripts/movies.csv")

# The FiveThirtyEight dataset normally has columns named:
# binary, budget, intgross
bechdel_roi <- bechdel |>
  filter(
    binary == "PASS",
    !is.na(budget),
    !is.na(intgross),
    budget > 0,
    intgross > 0
  ) |>
  mutate(
    intgross = suppressWarnings(as.numeric(intgross)),
    roi = intgross / budget
  )

# We again pretend that we observed only a random sample.
set.seed(2409)
bechdel_sample <- bechdel_roi |>
  slice_sample(n = min(50, nrow(bechdel_roi)))
```


## Before Calculating Anything...

What do you expect the distribution of ROI to look like?

. . .

- Symmetric?
- Left-skewed?
- Right-skewed?
- Any outliers?


## Look at the Raw Data

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(bechdel_sample, aes(x = roi)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 1, linetype = 2, linewidth = 1) +
  labs(
    x = "Worldwide gross / production budget",
    y = "Number of movies",
    title = "ROI among sampled movies that pass the Bechdel test"
  )
```

What do you notice?


## Mean or Median?

```{r}
#| include: false

movie_n <- nrow(bechdel_sample)
movie_mean <- mean(bechdel_sample$roi)
movie_median <- median(bechdel_sample$roi)
movie_sd <- sd(bechdel_sample$roi)
movie_se <- movie_sd / sqrt(movie_n)
```

::: {style="font-size: 105%;"}

- Mean ROI = `{r} round(movie_mean, 2)`
- Median ROI = `{r} round(movie_median, 2)`
- SD = `{r} round(movie_sd, 2)`

:::

. . .

Why can the mean and median differ substantially here?


## But Our Hypothesis Is About a Mean

Let $\mu$ be the **population mean ROI** among passing movies in our sampling frame.

We ask whether the average exceeds break-even:

$$
\hypo_0: \mu = 1
$$

against

$$
\hypo_1: \mu > 1.
$$

. . .

This is a **one-tailed** hypothesis.


## Your Turn: What Comes Next?

<!-- Give students 1-2 minutes before revealing. -->

We know:

- $\bar{x} = `{r} round(movie_mean, 2)`$
- $s = `{r} round(movie_sd, 2)`$
- $n = `{r} movie_n`$

What should we calculate next?

. . .

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}.
$$


## Standard Error

$$
SE_{\bar{x}}
=
\frac{`r round(movie_sd, 2)`}
{\sqrt{`r movie_n`}}
=
`r round(movie_se, 3)`.
$$

. . .

Notice:

> A highly variable raw distribution can still give a much less variable distribution of **sample means**.


## Your Turn: Test Statistic

Under $\hypo_0$, the expected population mean is 1.

What is the test statistic?

. . .

```{r}
#| include: false

movie_t <- (movie_mean - 1) / movie_se
movie_df <- movie_n - 1
movie_p <- 1 - pt(movie_t, df = movie_df)
```

$$
t
=
\frac{`r round(movie_mean, 2)` - 1}
     {`r round(movie_se, 3)`}
=
`r round(movie_t, 2)`.
$$


## Interpret It Before Looking at $p$

Our observed mean is **`{r} abs(round(movie_t, 2))` standard errors** above the null value.

. . .

Would you expect the $p$-value to be:

- large,
- around .05,
- or very small?


## Sampling Distribution and $p$-value

```{r}
#| fig-width: 9
#| fig-height: 5

movie_x_t <- seq(-4, max(6, movie_t + 1), length.out = 1200)
movie_t_plot <- data.frame(
  t = movie_x_t,
  density = dt(movie_x_t, df = movie_df)
)

ggplot(movie_t_plot, aes(t, density)) +
  geom_line(linewidth = 1.2) +
  geom_area(
    data = subset(movie_t_plot, t >= movie_t),
    aes(y = density),
    alpha = .35
  ) +
  geom_vline(xintercept = movie_t, linetype = 2) +
  labs(
    x = "Test statistic under H0",
    y = "Density",
    title = "One-tailed probability under the null hypothesis"
  )
```


## Conclusion

$$
p `{r} pretty_p(movie_p)`.
$$

Using $\alpha=.05$:

```{r}
#| results: asis
if (movie_p < .05) {
  cat("**We reject $\\mathcal{H}_0$.**")
} else {
  cat("**We do not reject $\\mathcal{H}_0$.**")
}
```

. . .

Complete this sentence:

> "If the null hypothesis were true, ..."


## Confidence Interval

```{r}
#| include: false

movie_crit <- qt(.975, df = movie_df)
movie_ci <- movie_mean + c(-1, 1) * movie_crit * movie_se
```

The ordinary two-sided 95% confidence interval is

$$
[`r round(movie_ci[1], 2)`,\; `r round(movie_ci[2], 2)`].
$$

What does this add beyond the $p$-value?


## Does Skewness Break Everything?

The **raw ROI values** are strongly skewed.

But our test concerns the **sample mean**.

So the relevant question is:

> What does the sampling distribution of the sample mean look like?


## Simulate the Sampling Distribution

```{r}
#| include: false

set.seed(2410)
B <- 5000

sim_movie_means <- replicate(
  B,
  mean(sample(bechdel_roi$roi, size = movie_n, replace = TRUE))
)

sim_movies <- data.frame(mean = sim_movie_means)
```


## Raw Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(bechdel_roi, aes(x = roi)) +
  geom_histogram(bins = 50) +
  labs(
    x = "ROI",
    y = "Number of movies",
    title = "Individual movies: strongly right-skewed"
  )
```


## Sampling Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(sim_movies, aes(x = mean)) +
  geom_histogram(bins = 40) +
  labs(
    x = paste0("Mean ROI in a sample of ", movie_n, " movies"),
    y = "Number of simulated samples",
    title = "Means of repeated samples"
  )
```

. . .

The Central Limit Theorem is about the **second** distribution.

## Sampling Distribution

```{r}
#| fig-width: 9
#| fig-height: 5

set.seed(2410)
B <- 5000

sim_movie_means_100 <- replicate(
  B,
  mean(sample(bechdel_roi$roi, size = 100, replace = TRUE))
)

sim_movies_100 <- data.frame(mean = sim_movie_means_100)

ggplot(sim_movies, aes(x = mean)) +
  geom_histogram(bins = 40) +
  labs(
    x = paste0("Mean ROI in a sample of ", movie_n, " movies"),
    y = "Number of simulated samples",
    title = "Means of repeated samples"
  )
```


## A More Important Problem

Suppose our calculations are all correct.

What exactly can we generalize to?

. . .

The dataset is **not** a simple random sample of every movie ever made.

The inferential calculation does not repair problems in the sampling frame.


## Statistical Significance Is Not Enough

A small $p$-value answers a narrow question:

> Are these data difficult to reconcile with $\hypo_0$?

It does **not** automatically tell us:

- whether the effect is important,
- whether the measurement is ideal,
- whether the sample represents the target population,
- whether one variable **causes** another.

<!-- TODO: Keep causality discussion brief here; full treatment comes with regression. -->


## Case Study 2: What Changed?

The statistical machinery was nearly identical.

But the data raised new questions about:

- skewness and outliers,
- mean versus median,
- operationalization,
- the target population,
- external validity.

. . .

Good statistics is more than calculating the correct formula.


# Case Study 3: Us {.section-slide}


## How Much Did This Room Sleep?

At the start of the lecture, you reported how many hours you slept last night.

Now we have a dataset.


## Load the Class Data

```{r}
#| include: false

# TODO: REPLACE THIS PLACEHOLDER VECTOR WITH THE LIVE RESPONSES.
#
# Easy option during preparation:
# sleep_hours <- c(7, 6.5, 8, ...)
#
# Or, if you export a polling tool:
# sleep_data <- read.csv("scripts/sleep_today.csv")
# sleep_hours <- sleep_data$hours
#
# The placeholder below exists only so the .qmd renders before class.

sleep_hours <- c(
  7.0, 6.5, 8.0, 7.5, 6.0, 7.0, 8.5, 5.5, 7.0, 6.0,
  7.5, 8.0, 6.5, 7.0, 9.0, 5.0, 6.5, 7.5, 8.0, 6.0,
  7.0, 7.0, 6.5, 8.5, 7.5, 5.5, 6.0, 7.0, 8.0, 6.5
)

sleep <- data.frame(hours = sleep_hours) |>
  filter(!is.na(hours), hours > 0, hours <= 24)
```


## The Raw Data

```{r}
#| fig-width: 9
#| fig-height: 5

ggplot(sleep, aes(x = hours)) +
  geom_histogram(binwidth = .5, boundary = 0) +
  geom_vline(xintercept = 8, linetype = 2, linewidth = 1) +
  labs(
    x = "Hours slept last night",
    y = "Number of students",
    title = "Sleep in this room"
  )
```


## Describe the Sample

```{r}
#| include: false

sleep_n <- nrow(sleep)
sleep_mean <- mean(sleep$hours)
sleep_median <- median(sleep$hours)
sleep_sd <- sd(sleep$hours)
sleep_se <- sleep_sd / sqrt(sleep_n)
```

- $n = `{r} sleep_n`$
- Mean = `{r} round(sleep_mean, 2)` hours
- Median = `{r} round(sleep_median, 2)` hours
- SD = `{r} round(sleep_sd, 2)` hours

. . .

Do **we** look sleep deprived?


## What Is the Hypothesis?

Suppose we choose eight hours as the benchmark.

If our claim was decided **before looking at the data**:

$$
\hypo_0: \mu = 8
$$

$$
\hypo_1: \mu < 8.
$$

. . .

Why does it matter whether we chose the direction before or after seeing the histogram?


## Discovery vs. Confirmation

Imagine we first inspect the data and notice that the mean is below 8.

Then we announce:

> "Our hypothesis was that students sleep **less** than eight hours."

. . .

That is not the same as having predicted the direction beforehand.

**Exploration can generate hypotheses. Confirmation tests hypotheses specified in advance.**


## You Know the Routine

Calculate:

$$
SE_{\bar{x}} = ?
$$

and

$$
t = ?
$$

```{r}
#| include: false

sleep_t <- (sleep_mean - 8) / sleep_se
sleep_df <- sleep_n - 1
sleep_p <- pt(sleep_t, df = sleep_df)
sleep_crit <- qt(.975, df = sleep_df)
sleep_ci <- sleep_mean + c(-1, 1) * sleep_crit * sleep_se
```


## Solution

$$
SE_{\bar{x}}
=
\frac{`r round(sleep_sd, 2)`}
{\sqrt{`r sleep_n`}}
=
`r round(sleep_se, 3)`.
$$

$$
t
=
\frac{`r round(sleep_mean, 2)` - 8}
{`r round(sleep_se, 3)`}
=
`r round(sleep_t, 2)`.
$$

$$
p `{r} pretty_p(sleep_p)`.
$$


## Confidence Interval

The 95% confidence interval is

$$
[`r round(sleep_ci[1], 2)`,\; `r round(sleep_ci[2], 2)`]
\text{ hours}.
$$

. . .

We now have a mathematically valid analysis.

But have we answered the question we thought we were answering?


## What Is Our Population?

Which of these is justified?

- Students in the Netherlands?
- Tilburg University students?
- Liberal Arts students?
- Students enrolled in this course?
- Students who attended today's lecture?
- Students who attended **and answered the poll**?

. . .

The statistical test does not decide this for us.


## Selection Matters

Today's sample might systematically exclude:

- students who skipped because they were tired,
- students who had other commitments,
- students unwilling to answer,
- students who attend another lecture group.

. . .

A tiny standard error cannot compensate for a badly defined or unrepresentative sample.


## Deduction and Induction

We used **deduction** when we said:

> If $\hypo_0$ is true, what should repeated samples look like?

We use **induction** when we move in the other direction:

> We observed this sample; what does it tell us about a larger population?

. . .

The second step always requires assumptions about how the data were obtained.


# Three Studies, One Logic {.section-slide}


## Same Machinery, Different Stories

| | Billboard | Bechdel | Sleep |
|---|---|---|---|
| Variable | Song length | ROI | Hours slept |
| Null value | 180 sec | 1 | 8 h |
| Alternative | $\neq$ | $>$ | $<$ |
| Main challenge | Sampling variability | Skew + measurement | Sampling / selection |
| Statistic | Mean | Mean | Mean |
| Core tool | SE + test statistic | SE + test statistic | SE + test statistic |


## The Core Pattern

In all three cases:

$$
\text{observed sample}
\longrightarrow
\text{estimate}
\longrightarrow
\text{standard error}
$$

$$
\longrightarrow
\text{sampling distribution under }\hypo_0
\longrightarrow
p
\longrightarrow
\text{conclusion}.
$$

The **story** changes.

The statistical logic does not.


## What Does a Test Statistic Mean?

Suppose

$$
t = -2.4.
$$

What does that mean?

. . .

> The observed estimate lies **2.4 standard errors below** the value predicted by the null hypothesis.

This interpretation is much more useful than memorizing formulas.


## What Does a $p$-value Mean?

Suppose $p=.03$.

Which statement is correct?

A. There is a 3% probability that $\hypo_0$ is true.

B. There is a 3% probability that the result occurred by chance.

C. If $\hypo_0$ were true, results at least this extreme would occur about 3% of the time.

D. There is a 97% probability that $\hypo_1$ is true.

. . .

**C**


## What Does a Confidence Interval Mean?

A 95% confidence interval is **not**:

> "There is a 95% probability that $\mu$ lies in this particular interval."

Instead:

> If we repeated the procedure many times, 95% of the intervals produced by this procedure would contain the true population value.


## Raw Distribution vs. Sampling Distribution

This distinction appeared repeatedly today.

**Raw-data distribution**

- describes individual observations
- e.g., individual movies or songs

**Sampling distribution**

- describes a statistic over repeated samples
- e.g., sample means
- determines the uncertainty of our estimate


## Standard Deviation vs. Standard Error

**Standard deviation**

> How much do individual observations vary?

$$
s
$$

**Standard error**

> How much would our sample estimate vary over repeated samples?

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}
$$


## What Happens When $n$ Increases?

If the population variability stays the same:

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}.
$$

As $n$ increases:

- the standard error decreases,
- the sampling distribution becomes narrower,
- estimates become more precise,
- power generally increases.


## But Large $n$ Does Not Fix Everything

A huge sample does **not** automatically fix:

- poor measurement,
- selection bias,
- an irrelevant target population,
- confounding,
- bad research questions.

. . .

Statistics quantifies **sampling uncertainty**.

It does not make every other research problem disappear.

