---
title: "Case Studies 2"
subtitle: "Lecture 7 - LAS"
author: "Don van den Bergh"
date: 2026-10-05
date-format: "DD-MM-YYYY"
format:
    dtslides-revealjs:
        theme: [../common_resources/css/tilburg-university.scss]
        logo: ../common_resources/logos/logo_tu_small.svg
        slide-number: c/t
        toc: true
        toc-depth: 1
        toc-title: "Today"
        embed-resources: false
---

## Today {.unlisted}

$$
\newcommand{\hypo}{\mathcal{H}}
$$



Two statistical tests using two datasets:

- Z-test
- $t$-test

These are mostly the names for tests we've already applied during the previous lectures.

The workflow is mostly the same.

# The Statistical Workflow {.section-slide}


## From a Question to a Conclusion

::: {style="font-size: 90%;"}

1. What is the _research question_?
2. What are the _population_, _sample_, _variables_, and _measurement levels_?
3. What do the _raw data_ look like?
4. Which _descriptive statistics_ are useful?
5. What are $\hypo_0$ and $\hypo_1$?
6. What would we expect because of _sampling variability_ if $\hypo_0$ were true?
7. How far is our estimate from $\hypo_0$ in _standard errors_?
8. What is the _$p$-value_?
9. What does a _confidence interval_ tell us?
10. What _can_ and what _can't_ we conclude?

:::


## The Recurring Test Statistic

For a sample mean,

$$
SE_{\bar{x}} = \frac{s}{\sqrt{n}}.
$$

The generic test statistic is

$$
\text{test statistic}
=
\frac{\text{observed estimate} - \text{value under }\hypo_0}
     {\text{standard error}}.
$$

. . .

Interpretation:

_How many standard errors away from the null hypothesis is our estimate?_



# Case study 1: Can we trust an AI detector? {.section-slide}

## Case study: Can we trust an AI detector?

Universities increasingly use AI detectors to identify work that may have been generated by tools such as ChatGPT.

- E.g., portfolio assignments.

But what does it actually mean when an AI detector flags a submission?

. . .

A 2023 *Scientific Reports* study by Ibrahim et al. tested GPTZero on real student answers and ChatGPT-generated answers across 32 university courses.

They found that GPTZero:

- incorrectly flagged _18% of human-written answers_ as AI-generated;
- failed to flag _32% of genuinely AI-generated answers_.

::: footer
Source: Ibrahim et al. (2023), *Scientific Reports*
<https://www.nature.com/articles/s41598-023-38964-3>
:::

## Detector performance

- $A$ = the submission was AI-generated;
- $H$ = the submission was human-written;
- $F$ = GPTZero flags the submission as AI-generated.

From the study,

$$
P(F\mid H)=0.18 \quad \mathrm{and} \quad P(\mathrm{not}\; F\mid A)=0.32.
$$

. . .

What is $P(F\mid A)$?

. . .

Since an AI-generated answer is either flagged or not flagged,
$$
P(F\mid A) = 1 - P(\mathrm{not}\; F\mid A) = 1 - 0.32 = 0.68
$$

## A flagged submission

Suppose GPTZero flags a student's submission.

We are interested in

$$
P(A\mid F),
$$

the probability that the submission was actually AI-generated given that it was flagged.

. . .

Do we already know this probability from

$$
P(F\mid A)=0.68?
$$

. . .

No. $P(A\mid F)\neq P(F\mid A)$.


## What information are we missing?

Bayes' theorem gives

$$
P(A\mid F)
=
\frac{
P(F\mid A)P(A)
}{
P(F)
}.
$$

We know:

$$
P(F)
=
P(F\mid A)P(A)
+
P(F\mid H)P(H).
$$

Therefore,

$$
P(A\mid F)
=
\frac{
P(F\mid A)P(A)
}{
P(F\mid A)P(A)
+
P(F\mid H)P(H)
}.
$$

## What information are we missing?

We know the detector performance.

But we do _not_ yet know

$$
P(A)
$$

the proportion of submitted work that is actually AI-generated.

## The importance of the base rate

Let $\pi=P(A)$.

Then $P(H)=1-\pi$.

Using the reported GPTZero performance,

$$
P(A\mid F)
=
\frac{
0.68\pi
}{
0.68\pi+0.18(1-\pi)
}.
$$

The probability that a flagged submission is actually AI-generated depends on how common AI-generated submissions are in the first place.

## What if AI use is uncommon?

Suppose only 10% of submissions are AI-generated, i.e., $P(A)=0.10$.

Then

$$
\begin{align}
P(A\mid F)
&=
\frac{
0.68(0.10)
}{
0.68(0.10)+0.18(0.90)
}\\
&= \frac{0.068}{0.230} \approx 0.296
\end{align}
$$

. . .

Although GPTZero correctly flags 68% of AI-generated work,
only about 30% of flagged submissions would actually be AI-generated.

## The dean's proposal

Suppose the dean of Tilburg University argues:

::: {.callout-note appearance="minimal" icon=false}
We should only treat a GPTZero flag as meaningful evidence of AI use if a flagged submission has at least a 50% probability of actually being AI-generated.
:::

. . .

Mathematically speaking, this means

$$
P(A\mid F)\geq 0.50.
$$

. . .

Based on the reported detector performance, what proportion of submitted work would have to be AI-generated for this statement to be true?

## Finding the critical base rate {.scrollable}

We need to solve for $\pi$:

$$
\frac{
0.68\pi
}{
0.68\pi+0.18(1-\pi)
}
=
0.50.
$$

::: {style="font-size: 75%;"}
$$
\begin{align}
\frac{0.68\pi}{0.68\pi+0.18(1-\pi)}
&=
0.50
&& \text{given} \\[6pt]
0.68\pi
&=
0.50\left[0.68\pi+0.18(1-\pi)\right]
&& \text{multiply by denominator} \\[6pt]
0.68\pi
&=
0.34\pi+0.09(1-\pi)
&& \text{multiply by }0.50 \\[6pt]
0.68\pi
&=
0.34\pi+0.09-0.09\pi
&& \text{expand} \\[6pt]
0.68\pi+0.09\pi-0.34\pi
&=
0.09
&& \text{collect }\pi \\[6pt]
0.43\pi
&=
0.09
&& \\[6pt]
\pi
&=
\frac{0.09}{0.43}
&& \text{divide by }0.43 \\[6pt]
\pi
&\approx
0.209.
&&
\end{align}
$$

:::

## Interpretation

At least about $20.9\%$ of all submissions would need to be AI-generated before a GPTZero flag is more likely to represent AI-generated work than human-written work.

. . .

This gives us a new question:

::: {.callout-note appearance="minimal" icon=false}
Is AI use among Tilburg University submissions actually above 20.9%?
:::

We can investigate this using a sample.

## A Tilburg University study

Suppose Tilburg University randomly audits $n=200$ student submissions.

Assume the audit can reliably determine whether each submission is AI-generated.

Of the 200 submissions, 50 are AI-generated.

Therefore, the _sample proportion_ is

$$
\hat p
=
\frac{50}{200}
=
0.25.
$$

## Is 25% enough?

The dean's criterion requires the population proportion to exceed

$$
p=0.209.
$$

Our sample estimate is
$$
\hat p=0.25.
$$

Since

$$
0.25>0.209,
$$

can we conclude that more than 20.9% of all submissions are AI-generated?

## Is 25% enough?

No!

The difference could be due to _sampling variability_.

## Hypotheses

Let $p$ denote the proportion of all relevant Tilburg University submissions that are AI-generated.

We test

$$
\hypo_0:p\leq0.209 \quad \mathrm{against}\quad \hypo_1:p>0.209.
$$

. . .

This is a _one-sided_ test.

. . .

Like last week, we perform the test at the boundary of the null hypothesis:

$$
p=0.209.
$$

## From proportions to means


Recall from week 2 (Bechdel example) that if we define

$$
x_i=
\begin{cases}
1, & \text{if submission }i\text{ is AI-generated},\\
0, & \text{if submission }i\text{ is human-written}.
\end{cases}
$$

then, the sample mean is $\bar x=\frac{1}{n}\sum_{i=1}^nx_i=\hat p.$
i.e., sample proportion of AI-generated submissions.

. . .

So testing a population _proportion_ is equivalent to testing the population _mean of a binary variable_.

. . .

What does this tell us about the _sampling distribution_?

. . .

It's normal because of the central limit theorem

::: {style="font-size: 80%;"}

There are corrected sampling distributions that incorporate that proportions are between 0 and 1.

:::

## Sampling distribution under the null

For a binary variable with probability $p$,

$$
\sigma_X
=
\sqrt{p(1-p)}.
$$

Under the null hypothesis, $p_0=0.209$ and therefore,

$$
\sigma_X
=
\sqrt{0.209(1-0.209)}.
$$

The standard error of the sample mean is

$$
SE
=
\sigma_X/\sqrt{n}
=
\sqrt{
\frac{0.209(1-0.209)}{200}
}\approx0.0287.
$$

::: notes
need to tell that the variance is not exam material
:::

## The Z statistic

We observed $\hat p=0.25$ while $\hypo_0$ stipulates: $p_0=0.209.$

Therefore,

$$
Z
=
\frac{
\hat p-p_0
}{
SE
}
=
\frac{
0.25-0.209
}{
0.0287
}
\approx1.43.
$$

Using the Z-table, what is the $p$-value?


## The p-value

For $Z\approx1.43,$ we obtain $P(Z>1.43) = 0.076.$

At $\alpha=0.05,$

we _do not reject_ the null hypothesis.

. . .

The observed proportion, 25%, is above the required 20.9%.

But the sample does not provide sufficiently strong evidence to conclude that the population proportion exceeds 20.9%.

## What if we had more data?

Suppose instead Tilburg University audits

$$
n=1000
$$

submissions.

Again, we find that $25\%$ of them are AI-generated.

So the observed proportion remains

$$
\hat p=0.25.
$$

. . .

What changes?

## A larger sample

The standard error is now

$$
SE
=
\frac{\sigma_X}{\sqrt{n}}
=
\frac{\sqrt{p_0(1-p_0)}}{\sqrt{n}}
=
\sqrt{
\frac{0.209(1-0.209)}{1000}
}
\approx0.0129.
$$

_Increasing_ the sample size made the standard error _smaller_.

. . .

Therefore,

$$
Z
=
\frac{
\hat p-p_0
}{
SE
}
=
\frac{
0.25-0.209
}{
0.0129
}
\approx3.18.
$$

. . .

Now, $p < 0.001$, and we would reject $\hypo_0$.

## Same effect, different evidence

In both studies,

$$
\hat p=0.25.
$$

But:

$$
\begin{align}
n=200
\qquad&\Rightarrow\qquad
Z\approx1.43\\
n=1000
\qquad&\Rightarrow\qquad
Z\approx3.18.
\end{align}
$$

- The estimated proportion did not change.
- The _uncertainty about that estimate_ changed.


## Recap

Our original question was

$$
P(A\mid F).
$$

Bayes' theorem showed that the answer depends on

1. how well the detector works, and
2. how common AI-generated submissions are.

Using the GPTZero results,

$$
P(A\mid F)
=
\frac{
0.68P(A)
}{
0.68P(A)+0.18P(H)
}.
$$

## Recap

The significance test addressed a different question:

Do the data provide evidence that AI use is common enough for a GPTZero flag to be more likely correct than incorrect?

<!-- ## One more complication...

The Ibrahim et al. study also investigated what happened when ChatGPT-generated answers were paraphrased using QuillBot.

For these answers, GPTZero failed to identify approximately **95%** of the AI-generated texts.

Thus,

$$
P(\overline F\mid A)=0.95,
$$

so

$$
P(F\mid A)=0.05.
$$

. . .

What happens to our conclusions? -->

<!-- ## The dean's criterion again

We now need

$$
\frac{
0.05\pi
}{
0.05\pi+0.18(1-\pi)
}
\geq0.50.
$$

The boundary occurs when

$$
0.05\pi
=
0.18(1-\pi).
$$

Therefore,

$$
0.23\pi=0.18,
$$

and

$$
\boxed{
\pi\approx0.783
}.
$$

So approximately **78.3% of submissions would need to be AI-generated** before a GPTZero flag was more likely to correspond to AI-generated than human-written work under these detector-performance numbers. -->

## What did we learn?

A statement like

- "GPTZero detects 68% of AI-generated submissions"

does _not_ tell us

- "A submission flagged by GPTZero has a 68% probability of being AI-generated."

To answer the second question, we also need the _base rate_.

## What did we learn?

Bayes' theorem combines

$$
P(F\mid A),
\qquad
P(F\mid H),
\qquad
P(A)
$$

to determine

$$
P(A\mid F).
$$

. . .

If we estimate the base rate from data, we must account for _sampling uncertainty_.

. . .

Note that this is very similar to how a $p$-value tells us how probable a test statistic at least this extreme would be if $\hypo_0$ were true, but does not tell us $P(\hypo_0\mid\text{data})$!

## Finally...

The detector results also come from a particular study:

- particular student assignments;
- particular university courses;
- a particular version of ChatGPT;
- a particular version of GPTZero;
- data collected in 2023.

. . .

If I apply this to your portfolio assignments, would these probabilities necessarily be the same?

. . .

Probably not.

# Case study 2: Do we know how much we use our smartphones?

## Case study: Do we know how much we use our smartphones?

How many hours did you use your smartphone yesterday?

. . .

And how many separate times did you pick it up or check it?

. . .

How accurate do you think your estimates would be?

## Self-report versus actual behaviour

Andrews et al. (2015) asked 23 participants to estimate their typical daily smartphone use.

At the same time, an app objectively recorded their smartphone behavior for two weeks.

For every participant we therefore have:

- estimated number of daily uses;
- actual number of daily uses;
- estimated daily duration;
- actual daily duration.

::: footer
Source: Andrews et al. (2015), *PLoS ONE*
<https://doi.org/10.1371/journal.pone.0139004>
:::

## The data

```{r}
#| echo: false

library(ggplot2)

phone <- read.csv("data/smartphone_use.csv")

phone$duration_difference <- phone$estimate_duration - phone$actual_duration
phone$uses_difference <- phone$estimated_uses - phone$actual_uses
```

Each row represents one participant.

```{r}
#| echo: false
#| output: asis

knitr::kable(
  head(phone[, c("estimated_uses", "actual_uses", "estimate_duration", "actual_duration")]),
  digits = 2,
  col.names = c("Estimated uses", "Actual uses", "Estimated hours", "Actual hours")
)
```

## How long do we use our phones?

Let's first consider total daily duration.

Participants estimated that they used their smartphones for, on average,

$$
4.12\text{ hours per day}.
$$

The objective measurements showed an average of

$$
5.06\text{ hours per day}.
$$

. . .

So participants underestimated their smartphone use by almost an hour.

Is that convincing evidence that people systematically underestimate their smartphone use?

## Looking at the individual participants

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

ggplot(phone, aes(x = actual_duration, y = estimate_duration)) +
  geom_point(size = 3, colour = "black", fill = "grey", shape = 21) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(x = "Actual use (hours per day)", y = "Estimated use (hours per day)") +
  theme_minimal(base_size = 24)
```

The dashed line represents a perfect agreement.

. . .

Points below the line are to participants who _underestimated_ their use.

Points above the line are to participants who _overestimated_ their use.

## Two measurements per participant

For each participant, we now have _two observations_:

$$
\text{estimated use}
\qquad\text{and}\qquad
\text{actual use}.
$$

These observations are not independent.

. . .

But we can turn them into one variable.

For participant $i$, define

$$
D_i =
\text{estimated duration}_i
-
\text{actual duration}_i.
$$

## The difference score

If somebody estimates their phone use perfectly, $D_i=0.$

If they underestimate their use, $D_i<0.$

If they overestimate their use, $D_i>0.$

. . .

Our original question can therefore be rewritten as:

"Is the population mean of these difference scores equal to zero?"

## The differences

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

ggplot(phone, aes(x = duration_difference)) +
  geom_histogram(
    bins = 8,
    boundary = 0
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    x = "Estimated − actual duration (hours)",
    y = "Number of participants"
  ) +
  theme_minimal(base_size = 24)
```

The sample mean difference is

$$
\bar D
=
4.12-5.06
=
-0.93\text{ hours}.
$$

## Our hypotheses

Let $\mu_D$ denote the population mean difference between estimated and actual smartphone use.

We test

$$
\hypo_0:\mu_D=0
\quad
\mathrm{against}
\quad
\hypo_1:\mu_D\neq0.
$$

. . .

Although every participant gave us two measurements, this is now simply a one-sample test of the difference scores.

. . .

Is this a Z or a $t$-test?

. . .

$t$-test

## The test statistic

The 23 differences have $\bar D=-0.93$ and $s_D\approx2.49.$

So,

$$
SE_{\bar D}
=
\frac{s_D}{\sqrt{n}}
=
\frac{2.49}{\sqrt{23}}
\approx0.52.
$$

The test statistic is

$$
t
=
\frac{\bar D-0}{SE_{\bar D}}
=
\frac{-0.93}{0.52}
\approx-1.80.
$$

## Is this statistically significant?

With $\alpha=0.05$, $df=22$, and $t = -1.80$ we examine the $t$-table.

. . .

We find $p(t > 1.717) = 0.05$. Since $1.717 < 1.80$ we know $p < 0.05$.

Can we reject $\hypo_0$?

. . .

No!

We have $p(t > 1.717) = 0.05$ and $p(t > 2.074) = 0.025$. Since $1.717 < 1.80 < 2.074$, we have that the actual $p$ value is between 0.05 and 0.025.

But this is a _two-sided_ test, and the above are _one-tailed_ $p$ values.

We need to multiply by 2.

Thus, the actual $p$ value is between 0.10 and 0.05.

## Confidence interval

The corresponding 95% confidence interval is
$$
[
\bar D - t_{97.5\%,\,22} \times SE_{\bar D},\;
\bar D + t_{97.5\%,\,22} \times SE_{\bar D}
]
$$


$$
[-2.01,\;0.14]\text{ hours}.
$$

Note that the interval contains 0.

(and yes, like last week you cannot compute this interval for a $t$-test yourself)

## What can we conclude?

Participants estimated 4.12 hours but actually used their phones for 5.06 hours.

The observed difference was almost one hour.

. . .

But, $p > 0.05$.

With only 23 participants, the data do not provide significant evidence that people under- or overestimate their total smartphone use.

## Does that mean people know their smartphone use well?

Not necessarily.

. . .

So far we have only considered _total duration_.

The researchers asked another question:

"How many separate times do you use your phone each day?"

This is a different operationalization!

## How often do we use our phones?

Participants estimated an average of about $37.20$ uses per day.

. . .

Their phones recorded an average of $84.68$ uses per day.

. . .

Actual use was more than _twice_ as frequent as participants estimated.

## Looking at the individual participants

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

ggplot(
  phone,
  aes(x = actual_uses, y = estimated_uses)
) +
  geom_point(size = 3) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(
    x = "Actual uses per day",
    y = "Estimated uses per day"
  ) +
  theme_minimal(base_size = 24)
```

## Difference scores again

Define $D_i=\text{estimated uses}_i - \text{actual uses}_i.$

Then $\bar D = 37.20-84.68 = -47.48.$

So participants _underestimated_ their smartphone use by about $47.5$ separate uses _per day_.

## The distribution of differences

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

ggplot(phone, aes(x = uses_difference)) +
  geom_histogram(
    bins = 8,
    boundary = 0
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    x = "Estimated − actual uses per day",
    y = "Number of participants"
  ) +
  theme_minimal(base_size = 24)
```

Can we explain this difference by sampling variability?

## Hypotheses

Again, let $\mu_D$ denote the population mean difference.

We test

$$
\hypo_0:\mu_D=0
\quad
\mathrm{against}
\quad
\hypo_1:\mu_D\neq0.
$$

The sample gives $\bar D=-47.48$ and $s_D\approx57.95$.

## The test statistic

The standard error is $SE_{\bar D} = \frac{s_D}{\sqrt{n}} = \frac{57.95}{\sqrt{23}} \approx12.09$.

The test statistic is $t = \frac{\bar D - 0}{12.09} = \frac{-47.48}{12.09} \approx-3.93.$

With $df=22$, we examine the $t$-table.

. . .

We find $p(t > 3.505) = 0.001$.

Again, this is a one-sided $p$-value, so since $3.505 < 3.93$ we can conclude that

$$
p < 0.002
$$

## Conclusion

We reject

$$
\hypo_0:\mu_D=0.
$$

The data provide strong evidence that participants systematically misestimate how often they use their smartphones.

The 95% confidence interval for the mean difference is approximately

$$
[-72.55,\;-22.42].
$$

<!--
In other words, the population mean underestimation is consistent with roughly 22 to 73 uses per day.
-->

## An interesting contrast

::: {style="font-size: 90%;"}

For total time:

$$
\begin{aligned}
\text{Estimated} &= 4.12\text{ hours}\\
\text{Actual} &= 5.06\text{ hours}\\
p &> 0.05
\end{aligned}
$$

We did not find sufficient evidence of systematic bias.

. . .

For number of uses:

$$
\begin{aligned}
\text{Estimated} &= 37.20\\
\text{Actual} &= 84.68\\
p &<0.002
\end{aligned}
$$

People dramatically underestimated how often they used their phones.

:::

## What did we learn?

When the same person is measured twice, we can define a difference:

$$
D_i=X_{i,1}-X_{i,2}.
$$

Then a comparison between two paired measurements becomes a familiar one-sample problem:

$$
\hypo_0:\mu_D=0.
$$

The operationalization can determine the outcome!

## Finally...

The data were collected in **2015**.

Smartphones, apps, notifications, and people's usage patterns have changed substantially since then.

Would you expect the numerical results to be the same for students today?

. . .

Probably not necessarily.